---
title: "Housing Market Analysis (Kaggle)"
author: "Angel Leon"
output: slidy_presentation
date: "DATA 2401"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Libraries to Load

all of these libraries will be used as some point in the project 
```{r}


library(broom)
library(readr)
library(dplyr)  
library(dbplyr)  
library(tidyverse)
library(ggplot2)
```


# Data Overview

Let's Plot the data and see what we're working with
```{r}
Housing_Market <- read_csv("./housing_price_dataset.csv" , show_col_types = F)
glimpse(Housing_Market)
# seeing what data this contains and summarizing the data
summary(Housing_Market)

```

# Why use this Data

When beginning this project using this data I had one main question on my mind; Can I predict the Housing Market for the next 30 years using Trends in this data.

To do that I had to start small asking smaller question in order to get an idea of how I can achieve this. Some of the questions were: 
Is there any pattern or trend in house prices based on the year they were built?
How has the average price changed over the years in different neighborhoods?
Are there any interactions between variables that have a significant impact on price?

# Trend Analysis

To start I wanted to see if there was any trends so I made two graphs showing the average prices over the years.
```{r}
# Trends Analysis
# 1. Trend in house prices based on the year they were built
# Grouping data by year built and calculating the average price for each year
yearly_price_trend <- Housing_Market %>%
  group_by(YearBuilt) %>%
  summarise(avg_price = mean(Price))

# Plotting trend in house prices over the years
ggplot(yearly_price_trend, aes(x = YearBuilt, y = avg_price)) +
  geom_line() +
  labs(title = "Trend in House Prices Over Years", x = "Year Built", y = "Average Price")

# 2. Plotting the same trend but this time using the different types of neighborhoods by grouping year built and neighborhoods
neighborhood_price_trend <- Housing_Market %>%
  group_by(YearBuilt, Neighborhood) %>%
  summarise(avg_price = mean(Price))

# Plotting the average price change in different neighborhoods over the years
ggplot(neighborhood_price_trend, aes(x = YearBuilt, y = avg_price, color = Neighborhood)) +
  geom_line() +facet_wrap(~ Neighborhood)
  labs(title = "Average Price Change in Different Neighborhoods Over Years", x = "Year Built", y = "Average Price", color = "Neighborhood") +
  theme(legend.position = "bottom")


```
# Outliers

I wanted to see if there were any outliers that would affect the average prices, and to do this I used Z-scores. 

```{r}
# Identifying Outliers

# Calculating z-scores for price to identify outliers
Housing_Market$Price_z <- scale(Housing_Market$Price)

# Defining threshold for identifying outliers (usually |z-score| > 3)
outliers <- Housing_Market[abs(Housing_Market$Price_z) > 3, ]

# Summary of outliers
summary(outliers)

# Using a Scatter plot to visualize Outliers
ggplot(Housing_Market, aes(x = SquareFeet, y = Price)) +
  geom_point(data = outliers, aes(color = "Outlier"), alpha = 0.5) +
  geom_point(aes(color = "Non-Outlier")) +
  labs(title = "Scatter plot of Square Feet vs. Price with Outliers Highlighted", x = "Square Feet", y = "Housing Price", color = "Outlier Status")



```


# Linear Regression Model

One of the best ways to predict future trends is to use a Linear Regression Model and the one used here was helpful in seeing which of the Houses features (Bedrooms, Square feet, Neighborhood, etc.) are most influential in determining the House Price

```{r}
# Using a Linear Regression Model in order to help me predict future Housing Prices by seeing which of these features are the most influential in determining price
model <- lm(Price ~ SquareFeet + Bedrooms + Bathrooms + YearBuilt + Neighborhood, data = Housing_Market)

# Summary of the regression model
summary(model)

# Feature Importance / Coefficients
coefficients <- tidy(model)
top_features <- coefficients %>%
  arrange(desc(abs(estimate))) %>%
  select(term, estimate)

# Printing the top influential features
print(top_features)

```

# Predicitng House Market
In order to predict the Housing Market I created a dataframe containing the future years and the average value of key features from the historical data.
With this "Future Data" set up we used the Linear Regression Model in order to predict Housing Prices for these years.

```{r}
model <- lm(Price ~ SquareFeet + Bedrooms + Bathrooms + YearBuilt + Neighborhood, data = Housing_Market)


# Generate future years (2020-2050)
future_years <- seq(2020, 2050)

# Creating a dataframe with future years and other predictors (SquareFeet, Bedrooms, Bathrooms, YearBuilt, Neighborhood)
future_data <- expand.grid(
  SquareFeet = mean(Housing_Market$SquareFeet), 
  Bedrooms = mean(Housing_Market$Bedrooms),     
  Bathrooms = mean(Housing_Market$Bathrooms),   
  YearBuilt = mean(Housing_Market$YearBuilt),   
  Neighborhood = unique(Housing_Market$Neighborhood),  
  Year = future_years
)

# Predict housing prices for future years using the regression model
predicted_prices <- predict(model, newdata = future_data)

# Plotting the predicted housing prices for future years
plot(future_years, predicted_prices[1:length(future_years)], type = "l", col = "red", xlab = "Year", ylab = "Predicted Housing Price", main = "Predicted Housing Prices (2020-2050)")


```

# What Happened
As you can see this isn't a very good graph it very predictable and very stagnant.

To combat this and have a more realistic graph I found the appreciation and depreciation rates.

```{r}
# Calculating the year-over-year percentage change
yearly_price_trend <- yearly_price_trend %>%
  arrange(YearBuilt) %>%
  mutate(
    percent_change = (avg_price - lag(avg_price)) / lag(avg_price) * 100
  )

# Removing the first row since its percent change will be NA
yearly_price_trend <- yearly_price_trend[-1,]

# Calculating the average appreciation and depreciation rates
appreciation_rates <- yearly_price_trend %>%
  filter(percent_change > 0) %>%
  summarise(average_appreciation = mean(percent_change, na.rm = TRUE))

depreciation_rates <- yearly_price_trend %>%
  filter(percent_change < 0) %>%
  summarise(average_depreciation = mean(percent_change, na.rm = TRUE))

# Display the rates
print(appreciation_rates)
print(depreciation_rates)


```
# Using these rates to form a more "Realisitic Graph"
After finding these rates I used the same code but this time using random percent change based on the appreaciation and depreciation rates.

```{r}
model <- lm(Price ~ SquareFeet + Bedrooms + Bathrooms + YearBuilt + Neighborhood, data = Housing_Market)

# Generating future years (2020-2050)
future_years <- seq(2020, 2050)

# Creating a dataframe with future years and other predictors (SquareFeet, Bedrooms, Bathrooms, YearBuilt, Neighborhood)
future_data <- expand.grid(
  SquareFeet = mean(Housing_Market$SquareFeet),  # Replace with appropriate mean value or a realistic value
  Bedrooms = mean(Housing_Market$Bedrooms),     # Replace with appropriate mean value or a realistic value
  Bathrooms = mean(Housing_Market$Bathrooms),   # Replace with appropriate mean value or a realistic value
  YearBuilt = mean(Housing_Market$YearBuilt),   # Replace with appropriate mean value or a realistic value
  Neighborhood = unique(Housing_Market$Neighborhood),  # Replace with an appropriate neighborhood category
  Year = future_years
)



# Using the Predict function to predict housing prices for future years using the regression model
predicted_prices <- predict(model, newdata = future_data)

# Using random percent change and using the appreciation and depreciation rates
min_percent_change <- -0.0124  # -1.24
max_percent_change <- 0.0154  # 1.54

# Generate random percentage changes for each year within the specified range
random_percent_changes <- runif(length(predicted_prices), min = min_percent_change, max = max_percent_change)

# Apply the random percentage changes to the predicted prices for each year
adjusted_predicted_prices <- predicted_prices

for (i in 2:length(adjusted_predicted_prices)) {
  adjusted_predicted_prices[i] <- adjusted_predicted_prices[i - 1] * (1 + random_percent_changes[i])
}

# Plotting predicted housing prices for future years
plot(future_years, predicted_prices[1:length(future_years)], type = "l", col = "red", xlab = "Year", ylab = "Predicted Housing Price", main = "Predicted Housing Prices (2020-2050)")

# Plotting adjusted predicted housing prices for future years
plot(future_years, adjusted_predicted_prices[1:length(future_years)], type = "l", col = "blue", xlab = "Year", ylab = "Predicted Housing Price", main = "Adjusted Predicted Housing Prices (2020-2050)")

ggplot(yearly_price_trend, aes(x = YearBuilt, y = avg_price)) +
  geom_line() +
  labs(title = "Trend in House Prices Over Years", x = "Year Built", y = "Average Price")





```

# Conclusion
While this new graph may be more "realistic" is worth noting that this is a projection and it is working based off trends in the past and assuming those trends continue.

